% Niniejszy plik stanowi przykład formatowania pracy magisterskiej na
% Wydziale MIM UW.  Szkielet użytych poleceń można wykorzystywać do
% woli, np. formatujac wlasna prace.
%
% Zawartosc merytoryczna stanowi oryginalnosiagniecie
% naukowosciowe Marcina Wolinskiego.  Wszelkie prawa zastrzeżone.
%
% Copyright (c) 2001 by Marcin Woliński <M.Wolinski@gust.org.pl>
% Poprawki spowodowane zmianami przepisów - Marcin Szczuka, 1.10.2004
% Poprawki spowodowane zmianami przepisow i ujednolicenie 
% - Seweryn Karłowicz, 05.05.2006
% Dodanie wielu autorów i tłumaczenia na angielski - Kuba Pochrybniak, 29.11.2016

% dodaj opcję [licencjacka] dla pracy licencjackiej
% dodaj opcję [en] dla wersji angielskiej (mogą być obie: [licencjacka,en])
\documentclass[licencjacka,en]{pracamgr}

\usepackage{times}
% \usepackage[marginparwidth=75pt]{geometry}
% Optional math commands from https://github.com/goodfeli/dlbook_notation.
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}
\usepackage{times}
% additional packages
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[capitalise]{cleveref}
\usepackage{wrapfig}
\usepackage{multirow}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularray}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{cleveref}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{fontawesome5}
\usepackage{natbib}  % or \usepackage{biblatex} if you prefer biblatex
\bibliographystyle{plainnat}  % or another style of your choice

\usepackage{xargs}                      % Use more than one optional

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\isb}{I\textsuperscript{2}SB}
\newcommand{\isbs}{I\textsuperscript{2}SBs}
\newcommand{\todo}[1]{\textcolor{black}{#1}}
\newcommand{\note}[1]{\textcolor{orange}{#1}}
\newcommand{\method}{EquiDiff}  % Equivariant Diffusion Sampling
\newcommand{\framework}{Invariants}  % Invariant Images
% DIVA (Diverse Invariant Visual Analysis)
% GUIDE (Generative Understanding of Invariant Domains via Equivariance)

\newcommand{\diff}{\mathrm{d}}
\newcommand{\scoret}{\nabla_{\mathbf{x}_t} \log{p(\mathbf{x}_t, t)}}
\newcommand{\scoren}{\nabla_{\mathbf{x}_n} \log{p(\mathbf{x}_n, n)}}
\newcommand{\scoretwithcond}{\nabla_{\mathbf{x}_t} \log{p(\mathbf{x}_t, t \mid \mathbf{y})}}
\newcommand{\scorenwithcond}{\nabla_{\mathbf{x}_n} \log{p(\mathbf{x}_n, n \mid \mathbf{y})}}
\newcommand{\scoretcond}{\nabla_{\mathbf{x}_t} \log{p(\mathbf{y} \mid \mathbf{x}_t, t)}}
\newcommand{\scorencond}{\nabla_{\mathbf{x}_n} \log{p(\mathbf{y} \mid \mathbf{x}_n, n)}}
\newcommand{\nonlscoret}{\nabla_{\mathbf{x}_t} \log \boldsymbol{\Psi}(\mathbf{x}_t, t)}
\newcommand{\nonlhscoret}{\nabla_{\mathbf{x}_t} \log \widehat{\boldsymbol{\Psi}}(\mathbf{x}_t, t)}
\newcommand{\boldpsi}{\boldsymbol{\Psi}}
\newcommand{\boldhatpsi}{\widehat{\boldsymbol{\Psi}}}
\newcommand{\boldpsistar}{\boldsymbol{\Psi}^*}
\newcommand{\boldhatpsistar}{\widehat{\boldsymbol{\Psi}}^*}

\DeclareFontFamily{U}{matha}{\hyphenchar\font45}
\DeclareFontShape{U}{matha}{m}{n}{
      <5> <6> <7> <8> <9> <10> gen * matha
      <10.95> matha10 <12> <14.4> <17.28> <20.74> <24.88> matha12
      }{}
\DeclareSymbolFont{matha}{U}{matha}{m}{n}

\DeclareMathSymbol{\Lt}{3}{matha}{"CE}
\DeclareMathSymbol{\Gt}{3}{matha}{"CF}

% \pdfminorversion=7
% \pdfobjcompresslevel=0

% Dane magistranta:
\autor{Bartłomiej Sadlej}{429589}

\title{EquiDiff: Generative Exploration of Neural Network Invariant Sets through Diffusion-Based Sampling}
\titlepl{EquiDiff: Generatywna Eksploracja Zbiorów Niezmienniczych Sieci Neuronowych przez Próbkowanie Oparte na Dyfuzji}

\kierunek{Machine Learning}

% Praca wykonana pod kierunkiem:
\opiekun{prof. dr hab. inż. Przemysław Biecek\\
  Wydział Matematyki Informatyki i Mechaniki\\
  }

% miesiąc i~rok:
\date{September 2025}

%Podać dziedzinę wg klasyfikacji Socrates-Erasmus:
\dziedzina{ 
11.4 Sztuczna inteligencja\\ 
11.3 Informatyka\\
11.1 Matematyka
}

%Klasyfikacja tematyczna wedlug ACM (informatyka)
\klasyfikacja{I. Computing Methodologies\\
  I.2 Artificial Intelligence\\
  I.2.6 Learning\\
  I.2.10 Vision and Scene Understanding\\
  I.4 Image Processing and Computer Vision\\
  I.4.8 Scene Analysis
}

% Słowa kluczowe:
\keywords{explainable AI, generative models, diffusion models, invariant sets, level sets, neural network interpretability, mechanistic interpretability, sparse autoencoders, visual explanations, counterfactual generation, score-based generative models}

% Tu jest dobre miejsce na Twoje własne makra i~środowiska:
\newtheorem{defi}{Definition}[section]

% koniec definicji

\begin{document}
\maketitle

%tu idzie streszczenie na strone poczatkowa
\begin{abstract}
Understanding the decision-making process of deep neural networks is an active area of research in machine learning. Current state-of-the-art methods focus on finding human-interpretable concepts or features that influence predictions in known data samples. However, this work argues that this approach is limited in its ability to provide a comprehensive understanding of model behavior due to vast unexplored regions of the data manifold, not present in investigated datasets, which can potentially lead to the same predictions. This work has three contributions. The first one is a paradigm shift from traditional explainable AI (XAI) methods, which find human-interpretable features in known data, to generative XAI methods that synthesize new samples. Secondly, this work introduces a new framework \framework for generative XAI and thirdly, it proposes an efficient diffusion-based method \method for exploring it. Evaluation of this method on popular models such as ResNet-50 or Sparse Autoencoders (SAE) highlights significant limitations of current XAI methods. The second contribution of this work is a new method for exploring the invariant sets of neural networks. The third contribution is a new method for exploring the invariant sets of neural networks.
\end{abstract}

\tableofcontents

\input{sections/introduction.tex}

\input{sections/related_work.tex}

\input{sections/method.tex}

\input{sections/experiments.tex}

\input{sections/conclusion_and_future_work.tex}

\input{sections/appendix.tex}

\bibliography{references}

\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% coding: latin-2
%%% End:
